---
title: "Deep Analysis of College Matriculation"
author: "Nile Dixon"
date: "2/6/2021"
output:
  html_document: default
  pdf_document: default
---
```{r}
library(descr)
library(MASS)
library(corrplot)
library(ggplot2)
library(knitr)

```

## Links To Datasets
TxHigherEdData http://www.txhighereddata.org/index.cfm?objectid=DE7B0990-D970-11E8-BB650050560100A9
TEA SAT Data https://tea.texas.gov/reports-and-data/school-performance/accountability-research/college-admissions-testing-sat-and-act
TEA AP Data https://tea.texas.gov/reports-and-data/school-performance/accountability-research/advanced-placement-and-international-baccalaureate
TEA Ethnicity Data https://rptsvr1.tea.texas.gov/adhocrpt/adste.html
TEA Special Population Data https://rptsvr1.tea.texas.gov/adhocrpt/adspr.html

```{r}
#LOAD DATA INTO MEMORY
ap.data <- read.csv('data/ap_data.csv')
demographic.data <- read.csv('data/demographic_data.csv')
matriculation.data <- read.csv('data/matriculation_data.csv')
sat.data <- read.csv('data/sat_data.csv')
special.pops.data <- read.csv('data/special_pops_data.csv')
school.types.data <- read.csv('data/school_type_data.csv')

```

## Merging Files Together


```{r}
#MERGE FILES TOGETHER
master.file <- merge(matriculation.data, ap.data, by.x = c('CAMPUS.ID','YEAR'), 
                     by.y = c('CAMPUS.ID','YEAR'), all.x = TRUE, all.y = FALSE)
master.file <- merge(master.file, demographic.data, by.x = c('CAMPUS.ID','YEAR'),
                     by.y = c('CAMPUS.ID','YEAR'), all.x = TRUE, all.y = FALSE)
master.file <- merge(master.file, sat.data, by.x = c('CAMPUS.ID','YEAR'),
                     by.y = c('CAMPUS.ID','YEAR'), all.x = TRUE, all.y = FALSE)
master.file <- merge(master.file, special.pops.data, by.x = c('CAMPUS.ID','YEAR'),
                     by.y = c('CAMPUS.ID','YEAR'), all.x = TRUE, all.y = FALSE)
master.file <- merge(master.file, school.types.data, by.x = c('DISTRICT.ID.y','YEAR'),
                     by.y = c('DISTRICT.ID','YEAR'), all.x = TRUE, all.y = FALSE)
```


```{r}
#REMOVE NAs
master.file <- master.file[which(!is.na(master.file$CAMPUS.ID)),]

#FIXING MISSING DATA
master.file$SAT.PARTICIPATION.RATE[which(is.na(master.file$SAT.PARTICIPATION.RATE))] <- 66.0
master.file$AP.PARTICIPATION.RATE[which(is.na(master.file$AP.PARTICIPATION.RATE))] <- 26.0
master.file$SAT.ERW[which(is.na(master.file$SAT.ERW))] <- 500
master.file$SAT.MATH[which(is.na(master.file$SAT.MATH))] <- 500
master.file$DESCRIPTION <- toupper(master.file$DESCRIPTION)
master.file$DESCRIPTION[which(is.na(master.file$DESCRIPTION))] <- "MAJOR URBAN"

#STRINGS TO INTEGERS
master.file$HIGH.SCHOOL.STUDENTS <- as.integer(master.file$HIGH.SCHOOL.STUDENTS)
master.file$NOT.FOUND <- as.integer(master.file$NOT.FOUND)
master.file$NOT.TRACKABLE <- as.integer(master.file$NOT.TRACKABLE)
```


## Derive Statistics From Raw Data
```{r}
master.file$POTENTIAL.MATRICULANTS <- master.file$HIGH.SCHOOL.STUDENTS - master.file$NOT.TRACKABLE
master.file$MATRICULANTS <- master.file$POTENTIAL.MATRICULANTS - master.file$NOT.FOUND

master.file$TOTAL.DEMOGS <- master.file$ASIAN + master.file$NATIVE.AMERICAN + master.file$BLACK + master.file$HISPANIC + master.file$HAWAIIAN + master.file$TWO.RACES + master.file$WHITE

master.file$PERCENT.BLACK.BROWN <- (master.file$BLACK + master.file$HISPANIC) / master.file$TOTAL.DEMOGS
master.file$PERCENT.WHITE <- master.file$WHITE / master.file$TOTAL.DEMOGS
master.file$PERCENT.GT <- master.file$GT / master.file$ENROLLMENT
master.file$PERCENT.CTE <- master.file$CTE / master.file$ENROLLMENT
master.file$PERCENT.LEP <- master.file$LEP / master.file$ENROLLMENT
master.file$PERCENT.BIL <- master.file$BIL / master.file$ENROLLMENT
master.file$PERCENT.ESL <- master.file$ESL / master.file$ENROLLMENT
master.file$PERCENT.DISADV <- master.file$DISADV / master.file$ENROLLMENT
master.file$PERCENT.SPED <- master.file$SPED / master.file$ENROLLMENT
master.file$MATRICULATION.RATE <- master.file$MATRICULANTS / master.file$POTENTIAL.MATRICULANTS
master.file$SAT.PARTICIPATION.RATE <- master.file$SAT.PARTICIPATION.RATE / 100.0
master.file$AP.PARTICIPATION.RATE <- master.file$AP.PARTICIPATION.RATE / 100.0
master.file <- master.file[which(!is.na(master.file$MATRICULATION.RATE)),]
```

## Correlation Matrix
```{r}
master.numeric.df <- master.file[c('AP.PARTICIPATION.RATE','SAT.PARTICIPATION.RATE','SAT.MATH','SAT.ERW','PERCENT.BLACK.BROWN','PERCENT.WHITE','PERCENT.GT','PERCENT.CTE','PERCENT.LEP','PERCENT.BIL','PERCENT.ESL','PERCENT.DISADV','PERCENT.SPED','MATRICULATION.RATE')]
M <- cor(master.numeric.df, use = 'complete.obs')
corrplot(M, method = 'pie')
```

Based on the correlation matrix, these are the factors that seem to positively correlate with college matriculation:
1. Percent Took AP Exam
2. Percent Took SAT
3. Median SAT ERW
4. Median SAT Math
5. Percent Gifted and Talented
6. Percent Participated in Career and Technical Education

Based on the correlation matrix, these are the factors that seem to negatively correlate with college matriculation:
1. Percent Bilingual
2. Percent Limited English Proficiency (LEP)
3. Percent English as Second Language (ESL)
4. Percent Classified as Disadvantaged
5. Percent in Special Education


## Negative Binomial Regression
```{r}
base.model <- glm(MATRICULANTS / POTENTIAL.MATRICULANTS ~ 1, data = master.file)
model.bil <- glm(MATRICULANTS / POTENTIAL.MATRICULANTS ~ AP.PARTICIPATION.RATE + SAT.PARTICIPATION.RATE + SAT.MATH + SAT.ERW + PERCENT.BLACK.BROWN + PERCENT.WHITE + PERCENT.GT + PERCENT.BIL + PERCENT.DISADV + PERCENT.SPED + DESCRIPTION, data = master.file, family = "binomial")
model.esl <- glm(MATRICULANTS / POTENTIAL.MATRICULANTS ~ AP.PARTICIPATION.RATE + SAT.PARTICIPATION.RATE + SAT.MATH + SAT.ERW + PERCENT.BLACK.BROWN + PERCENT.WHITE + PERCENT.GT + PERCENT.ESL + PERCENT.DISADV + PERCENT.SPED + DESCRIPTION, data = master.file, family = "binomial")
model.lep <- glm(MATRICULANTS / POTENTIAL.MATRICULANTS ~ AP.PARTICIPATION.RATE + SAT.PARTICIPATION.RATE + SAT.MATH + SAT.ERW + PERCENT.BLACK.BROWN + PERCENT.WHITE + PERCENT.GT + PERCENT.LEP + PERCENT.DISADV + PERCENT.SPED + DESCRIPTION, data = master.file, family = "binomial")
summary(model.bil)
summary(model.esl)
summary(model.lep)
```



```{r}
campuses.2018 <- master.file[which(master.file$YEAR == 2018),]
campuses.2018$PREDICTED <- predict(model.bil, campuses.2018, type = 'response')
campuses.2018$DIFFERENCE <- campuses.2018$MATRICULATION.RATE - campuses.2018$PREDICTED

kable(campuses.2018[c('CAMPUS.NAME.y','PREDICTED','MATRICULATION.RATE')])
```